\section{Introduction}
\label{sec:introduction}

Self-supervised learning (SSL) methods based on joint embedding---DINO \citep{caron2021emerging}, DINOv2 \citep{oquab2023dinov2}, BYOL \citep{grill2020byol}, VICReg \citep{bardes2022vicreg}, and more recently SimDINO \citep{wu2025simdino}---learn representations that transfer to downstream tasks without labels. Most such methods $\ell_2$-normalize their output, so the learned features live on a unit hypersphere $\Sph^{d-1}$ and all feature norms are, by construction, equal to one. This choice is standard, well-motivated for contrastive losses, and usually not revisited.

Recent work on coding-rate regularization \citep{yu2020mcr2, wu2025simdino} and variance-covariance regularization \citep{bardes2022vicreg, zbontar2021barlow} makes it possible to train joint-embedding models \emph{without} the final $\ell_2$-normalization, instead constraining the distribution of features to be approximately isotropic Gaussian in $\R^d$. We refer to such models as \emph{flat-space} SSL, in contrast to the spherical-space SSL of DINO. In a flat-space model, feature norms are allowed to vary, so the magnitude $\|\vect{z}\|$ becomes an additional degree of freedom the network can use.

In this paper we ask a simple empirical question: \emph{what does the feature norm of a flat-space SSL model encode?} We train a ResNet-18 \citep{he2016resnet} on CIFAR-10 \citep{krizhevsky2009cifar} with a minimalist flat-Gaussian objective---a cross-view mean-squared alignment loss, a correlation-rate regularizer \citep{ma2007segmentation, wu2025simdino}, and a two-sided variance penalty \citep{schaeffer2024sigreg}---and analyze the geometry of the learned features. The findings are:

\begin{enumerate}
    \item \textbf{Feature norm tracks confidence}. The $k$-nearest-neighbour classification accuracy of a test sample is a strong monotone function of its feature norm: $44\%$ accuracy for the lowest-norm tenth of test samples, $92\%$ for the highest (\secref{sec:experiments}). The Pearson correlation between feature norm and distance from the empirical class centroid is $r = 0.99$.
    \item \textbf{Corruption contracts embeddings radially toward the origin}. Linearly mixing a real image with uniform random noise (mix fraction $\alpha \in [0, 1]$) monotonically reduces the feature norm from its clean value of $\sim 14.8$ to a stable floor of $\sim 11.5$ at $\alpha \ge 0.5$. The direction of the feature also drifts, but the radial contraction dominates.
    \item \textbf{Pure noise collapses onto a low-dimensional subspace orthogonal to the class manifold}. Real-data features are effectively full-rank ($\sim 237$ of $256$ dimensions), while pure-noise features concentrate in a $\sim 12$-dimensional subspace oriented at approximately $80^\circ$ from every class centroid. The encoder has, without supervision, carved out a distinct ``I do not know'' region.
    \item \textbf{The phenomenon does not appear in spherical-SSL models}. Running the same noise-injection experiment on the publicly available DINOv2 \citep{oquab2023dinov2} produces the opposite behaviour: the (unnormalized) CLS-token and patch-pooled feature norms \emph{increase} with noise fraction. Whatever is happening in our model is not a generic property of joint-embedding SSL; it is specific to the flat-Gaussian geometry.
\end{enumerate}

Findings in isolation have precedent. Concurrent work by Draganov et al.~\citep{draganov2025embedding_norms} establishes a norm--accuracy ranking in \emph{cosine-similarity} SSL (SimCLR, SimSiam, BYOL), attributing it to a density-of-updates mechanism on the pre-projection features; CSI \citep{tack2020csi} uses norm as an OOD score on SimCLR. Feature norm as a quality signal is familiar from face recognition \citep{meng2021magface, kim2022adaface}; feature/logit magnitude as an OOD score is the basis of energy-based detectors \citep{liu2020energy}; and dimensional collapse in contrastive SSL has been studied as a failure mode \citep{jing2022dimensional}. Our scope is specifically flat-space SSL---a setting not covered by \citep{draganov2025embedding_norms} and where the norm enters the training objective directly---and our contribution is to characterize the radial-contraction response to continuous input corruption, the low-dimensional pure-noise subspace, and a DINOv2 counterexample showing the sign of the coupling is not automatic.

The paper is organized as follows. \secref{sec:related_work} surveys the three literatures that converge here---flat-space SSL, feature-norm-as-quality, and OOD detection. \secref{sec:method} describes our training objective. \secref{sec:experiments} presents the empirical analysis, including the DINOv2 comparison.
